\documentclass[11pt]{article}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage[a4paper,margin=1in]{geometry}
\usepackage{hyperref}

\newcommand{\E}{\mathcal E}
\newcommand{\Var}{\mathsf{Var}}
\newcommand{\B}{\mathsf{B}}
\newcommand{\Tel}{\mathsf{T}}
\newcommand{\Ty}{\mathsf{Ar}}
\newcommand{\arity}{\mathsf{ar}}
\newcommand{\coh}{\mathsf{coh}}
\newcommand{\ext}{\mathbin{\triangleleft}}
\newcommand{\into}{\longrightarrow}
\newcommand{\id}{\mathrm{id}}
\newcommand{\pcomp}[3]{#1 \overset{#2}{\circ} #3}

\title{The Mathematical Content of \texttt{diamondAt}}
\author{}
\date{}

\begin{document}
\maketitle

\section*{The point}

The theorem \texttt{diamondAt} is an interchange law for hereditary
substitution.  It says that the following two operations commute:

\begin{enumerate}
\item replace variables from one context \(D\) by expressions over \(D'\);
\item replace variables from another context \(S\) by expressions over \(S'\).
\end{enumerate}

There is one important subtlety.  The fillers used to replace the \(S\)-variables
may themselves contain variables from \(D\).  Therefore, if we first substitute
\(D\)-variables, then the \(S\)-fillers must also be transformed by that same
substitution.  The theorem says that this is the only correction needed.

\section*{Notation}

We write \(\E(\Gamma)\) for expressions in context \(\Gamma\).  If \(\alpha\) is
an arity, then
\[
  \Gamma \ext \alpha
\]
means the context \(\Gamma\) extended by one binder of arity \(\alpha\).  This is
written \(\Gamma \mathbin{\texttt{::}} \alpha\) in Lean.

The operation \(\bowtie\) appends telescopes.  We use the following names for
the contexts appearing in the theorem:
\[
  P = pre,\qquad
  D = dom,\qquad
  D' = cod,\qquad
  T = \tau,\qquad
  S = src,\qquad
  S' = dst.
\]
The list \(\upsilon\) of already-opened binders gives a telescope
\[
  U = \Tel(\upsilon).
\]
In Lean, \(U\) is \(\mathsf{Tele.ofList}\,\upsilon\).

The expression to which the theorem applies lives in the context
\[
  P \bowtie D \bowtie T \bowtie S \bowtie U.
\]
Thus it may mention variables from five regions:
\[
  P,\qquad D,\qquad T,\qquad S,\qquad U.
\]

An argument family can be viewed as a singleton substitution.  In Lean, such
a family has type \texttt{Expr.Args Omega beta}.  If \(A\) is an argument
family for a head of arity \(\beta\), write
\[
  A^\sharp : \langle\beta\rangle \leadsto \Omega
\]
for the substitution defined by
\[
  A^\sharp(\mathsf{here}(j)) = A_j.
\]
This is Lean's \texttt{Subst.fromArgs beta Omega A}; conversely,
\texttt{Subst.toArgs} reads a singleton-domain substitution as its argument
family.  The beta-like step is then not a new primitive operation: it is
ordinary substitution action by \(A^\sharp\).
Thus
\[
  (A^\sharp)_\Xi(b)
\]
means: act on the filler \(b\) by the singleton substitution generated from the
argument family \(A\), at passive depth \(\Xi\).

\section*{The two substitutions}

The first substitution is
\[
  \sigma : D \leadsto P \bowtie D'.
\]
This means that for every variable \(z : D \ni \beta\), we have a filler
\[
  \sigma(z) \in \E(P \bowtie D' \ext \beta).
\]
It replaces \(D\)-variables while preserving the prefix \(P\).

The second substitution is
\[
  \kappa : S \leadsto P \bowtie D \bowtie T \bowtie S'.
\]
Thus for every \(x : S \ni \beta\),
\[
  \kappa(x)
  \in
  \E(P \bowtie D \bowtie T \bowtie S' \ext \beta).
\]
This replaces \(S\)-variables by expressions that still live over \(D\).  So
\(\kappa\) is a substitution before the change of target context induced by
\(\sigma\).

\section*{Action at a passive depth}

If \(A\) is a passive telescope, then \(\sigma\) acts at depth \(A\) as a map
\[
  \sigma_A :
  \E(P \bowtie D \bowtie A)
  \into
  \E(P \bowtie D' \bowtie A).
\]
The word ``passive'' means that variables from \(A\) are not substituted.  They
are merely carried along.  Only variables from \(D\) are active for \(\sigma\).

Similarly, \(\kappa\) acts at passive depth \(U\) as a map
\[
  \kappa_U :
  \E(P \bowtie D \bowtie T \bowtie S \bowtie U)
  \into
  \E(P \bowtie D \bowtie T \bowtie S' \bowtie U).
\]
Only variables from \(S\) are active for \(\kappa\); everything else is passive.

\section*{Postcomposition of a substitution}

Since the fillers of \(\kappa\) may contain \(D\)-variables, applying \(\sigma\)
to an expression also transforms the substitution \(\kappa\).  We write this
postcomposition as
\[
  \pcomp{\sigma}{T \bowtie S'}{\kappa} :
  S \leadsto P \bowtie D' \bowtie T \bowtie S'.
\]
It is defined pointwise by
\[
  (\pcomp{\sigma}{T \bowtie S'}{\kappa})(x)
  =
  \sigma_{T \bowtie S' \ext \beta}(\kappa(x)),
  \qquad x : S \ni \beta.
\]
The superscript \(T \bowtie S'\) records the passive depth at which
\(\sigma\) postcomposes \(\kappa\).  In the pointwise definition the full
passive depth is \(T \bowtie S' \ext \beta\), because the filler
\(\kappa(x)\) is an expression over
\[
  P \bowtie D \bowtie T \bowtie S' \ext \beta.
\]
After \(\sigma\) acts, this becomes an expression over
\[
  P \bowtie D' \bowtie T \bowtie S' \ext \beta,
\]
which is exactly the required type of a filler for \(x\) in the transformed
substitution.

\section*{The theorem as an equation}

Let
\[
  e \in \E(P \bowtie D \bowtie T \bowtie S \bowtie U).
\]
Then \texttt{diamondAt} says:
\[
\boxed{
  (\pcomp{\sigma}{T \bowtie S'}{\kappa})_U
  \bigl(
    \sigma_{T \bowtie S \bowtie U}(e)
  \bigr)
  =
  \sigma_{T \bowtie S' \bowtie U}
  \bigl(
    \kappa_U(e)
  \bigr).
}
\]

This is the clean mathematical statement.  The left-hand side first applies
\(\sigma\) to \(e\), then substitutes the remaining \(S\)-variables using the
postcomposition \(\pcomp{\sigma}{T \bowtie S'}{\kappa}\).  The right-hand
side first substitutes the \(S\)-variables using \(\kappa\), then applies
\(\sigma\).

\section*{The same equation as a square}

The theorem says that the following square commutes:
\[
\begin{array}{ccc}
\E(P \bowtie D \bowtie T \bowtie S \bowtie U)
& \xrightarrow{\ \sigma_{T \bowtie S \bowtie U}\ } &
\E(P \bowtie D' \bowtie T \bowtie S \bowtie U)
\\[0.8em]
\downarrow{\kappa_U} && \downarrow{(\pcomp{\sigma}{T \bowtie S'}{\kappa})_U}
\\[0.8em]
\E(P \bowtie D \bowtie T \bowtie S' \bowtie U)
& \xrightarrow{\ \sigma_{T \bowtie S' \bowtie U}\ } &
\E(P \bowtie D' \bowtie T \bowtie S' \bowtie U).
\end{array}
\]

The top arrow changes \(D\) into \(D'\).  The left arrow changes \(S\) into
\(S'\).  The bottom arrow changes \(D\) into \(D'\) after \(S\) has already
been changed into \(S'\).  The right arrow changes \(S\) into \(S'\) after the
fillers for \(S\) have themselves been postcomposed by \(\sigma\) at passive
depth \(T \bowtie S'\).

\section*{Why the theorem is not trivial}

If the fillers of \(\kappa\) did not contain variables from \(D\), the theorem
would be close to ordinary commutation of independent substitutions.  But here
the fillers of \(\kappa\) are expressions over
\[
  P \bowtie D \bowtie T \bowtie S',
\]
so they may contain \(D\)-variables.  Therefore substituting \(D\)-variables
before substituting \(S\)-variables changes the \(S\)-fillers themselves.

The postcomposition \(\pcomp{\sigma}{T \bowtie S'}{\kappa}\) is precisely the
bookkeeping that accounts for this dependency.  The theorem says:
\[
  \text{the dependency of \(\kappa\) on \(D\) is handled pointwise by \(\sigma\).}
\]

\section*{The five head cases}

The context of \(e\) decomposes into five regions:
\[
  P \bowtie D \bowtie T \bowtie S \bowtie U.
\]
Consequently, the head variable of an application in \(e\) belongs to exactly
one of:
\[
  U,\qquad S,\qquad T,\qquad D,\qquad P.
\]
The proof of the theorem is a structural recursion on \(e\), with a case split
on these five possibilities.

\paragraph{Head in \(U\).}
This is a local binder introduced by recursive descent.  Neither substitution
acts on it.  Both sides rebuild the same head and use the induction hypothesis
on all arguments.

More explicitly, suppose the expression has the form
\[
  e = \operatorname{ap}(x_\upsilon,a),
\]
where \(x_\upsilon : U \ni \beta\) is the head variable and, for each binder
argument \(j \in \B(\beta)\), the expression \(a(j)\) is the corresponding
subterm under the extra binder of arity \(\arity(j)\).

After unfolding the two routes in the diamond, both sides still have the same
head \(x_\upsilon\).  What changes is only the family of arguments.  For each
argument \(j\), define
\[
  L_j =
  (\pcomp{\sigma}{T \bowtie S'}{\kappa})_{U \ext \arity(j)}
  \bigl(
    \sigma_{T \bowtie S \bowtie U \ext \arity(j)}(a(j))
  \bigr)
\]
and
\[
  R_j =
  \sigma_{T \bowtie S' \bowtie U \ext \arity(j)}
  \bigl(
    \kappa_{U \ext \arity(j)}(a(j))
  \bigr).
\]
The induction hypothesis applied to the subterm \(a(j)\) says precisely that
\[
  L_j = R_j
\]
for every binder argument \(j\).  More precisely, for each \(j\) we call the
same theorem with the same \(P,D,D',T,S,S',\sigma,\kappa\), but with the
under-telescope enlarged from \(U\) to
\[
  U_j = U \ext \arity(j),
\]
and with expression \(a(j)\).  In Lean, using ASCII names for readability, this
is the recursive call
\begin{quote}
\small\ttfamily
diamondAt (upsilon := j.arity :: upsilon)\\
\hspace*{2em}sigma hTauSrc hTauDst srcTarget dstTarget kappa (args j)
\end{quote}
Therefore the under case concludes by
congruence of the application constructor and function extensionality:
\[
  \operatorname{ap}(x_\upsilon, j \mapsto L_j)
  =
  \operatorname{ap}(x_\upsilon, j \mapsto R_j).
\]
So the informal phrase ``the under case is
\(\operatorname{ap}(x_\upsilon, j \mapsto \text{induction hypothesis})\)''
means: after both routes have been normalized to applications with the same
head, the desired equality is induced pointwise from the recursive equalities
on the argument family.  In the Lean file, this argument is packaged by the
lemma \texttt{DiamondSite.underBranch}.

\paragraph{Head in \(S\).}
This is an active variable for \(\kappa\).  On the right side, \(\kappa\) fires
immediately.  On the left side, \(\sigma\) first acts on the ambient expression,
and then the postcomposition \(\pcomp{\sigma}{T \bowtie S'}{\kappa}\) fires.
The two results are equal because \(\pcomp{\sigma}{T \bowtie S'}{\kappa}\)
was defined by applying \(\sigma\) to the fillers of \(\kappa\).  This branch
recursively invokes the same theorem on the selected filler \(\kappa(x)\).

We now spell this out slowly, with the types visible.  Suppose
\[
  e = \operatorname{ap}(x_S,a),
  \qquad x_S : S \ni \beta.
\]
The symbol \(j\) below always ranges over the binders of the arity \(\beta\):
\[
  j : \B(\beta).
\]
The argument family \(a\) therefore has components
\[
  a(j) \in
  \E(P \bowtie D \bowtie T \bowtie S \bowtie U \ext \arity(j)).
\]
Write
\[
  U_j = U \ext \arity(j).
\]
Thus \(a(j) \in \E(P \bowtie D \bowtie T \bowtie S \bowtie U_j)\).

For readability, set
\[
  \rho = \pcomp{\sigma}{T \bowtie S'}{\kappa}
  :
  S \leadsto P \bowtie D' \bowtie T \bowtie S'.
\]
This is the postcomposition of \(\kappa\) by \(\sigma\).  For a slot
\(x : S \ni \gamma\), its filler is
\[
  \rho(x)
  =
  \sigma_{T \bowtie S' \ext \gamma}(\kappa(x))
  \in
  \E(P \bowtie D' \bowtie T \bowtie S' \ext \gamma).
\]
In particular,
\[
  \kappa(x_S)
  \in
  \E(P \bowtie D \bowtie T \bowtie S' \ext \beta),
\]
and
\[
  \rho(x_S)
  =
  \sigma_{T \bowtie S' \ext \beta}(\kappa(x_S))
  \in
  \E(P \bowtie D' \bowtie T \bowtie S' \ext \beta).
\]

With this notation, the branch goal is exactly
\[
  \rho_U
  \bigl(
    \sigma_{T \bowtie S \bowtie U}
      (\operatorname{ap}(x_S,a))
  \bigr)
  =
  \sigma_{T \bowtie S' \bowtie U}
  \bigl(
    \kappa_U(\operatorname{ap}(x_S,a))
  \bigr).
\]
Both sides live in
\[
  \E(P \bowtie D' \bowtie T \bowtie S' \bowtie U).
\]

First unfold the left-hand side.  The head \(x_S\) is passive for \(\sigma\),
because \(\sigma\) only replaces variables from \(D\).  Therefore \(\sigma\)
preserves the head and acts on the arguments:
\[
  \sigma_{T \bowtie S \bowtie U}(\operatorname{ap}(x_S,a))
  =
  \operatorname{ap}
  \Bigl(
    x_S,
    j \mapsto
      \sigma_{T \bowtie S \bowtie U \ext \arity(j)}(a(j))
  \Bigr).
\]
Write
\[
  \bar a_j =
  \sigma_{T \bowtie S \bowtie U_j}(a(j)).
\]
This is well-typed because
\[
  \sigma_{T \bowtie S \bowtie U_j} :
  \E(P \bowtie D \bowtie T \bowtie S \bowtie U_j)
  \to
  \E(P \bowtie D' \bowtie T \bowtie S \bowtie U_j),
\]
and \(a(j)\) is in the domain.  Hence
\[
  \bar a_j
  \in
  \E(P \bowtie D' \bowtie T \bowtie S \bowtie U_j).
\]
Then the previous equation says
\[
  \sigma_{T \bowtie S \bowtie U}(\operatorname{ap}(x_S,a))
  =
  \operatorname{ap}(x_S, j \mapsto \bar a_j).
\]

Now apply \(\rho_U\).  This substitution acts on \(x_S\), so it fires.  Firing
means: take the filler \(\rho(x_S)\) and substitute into its freshly opened
\(\beta\)-binder using the transformed argument family.  Define
\[
  L_j =
  \rho_{U_j}
  \bigl(
    \bar a_j
  \bigr).
\]
This is well-typed because
\[
  \rho_{U_j} :
  \E(P \bowtie D' \bowtie T \bowtie S \bowtie U_j)
  \to
  \E(P \bowtie D' \bowtie T \bowtie S' \bowtie U_j),
\]
and \(\bar a_j\) is in the domain.  Therefore
\[
  L_j
  \in
  \E(P \bowtie D' \bowtie T \bowtie S' \bowtie U_j).
\]
The family \(j \mapsto L_j\) gives a singleton-domain substitution
\[
  L^\sharp :
  \langle\beta\rangle
  \leadsto
  P \bowtie D' \bowtie T \bowtie S' \bowtie U.
\]
Its action at empty passive depth has type
\[
  (L^\sharp)_\varnothing :
  \E(P \bowtie D' \bowtie T \bowtie S' \ext \beta)
  \to
  \E(P \bowtie D' \bowtie T \bowtie S' \bowtie U).
\]
The input \(\rho(x_S)\) has exactly the required domain type
\(\E(P \bowtie D' \bowtie T \bowtie S' \ext \beta)\).  Hence
\[
  (L^\sharp)_\varnothing(\rho(x_S))
  \in
  \E(P \bowtie D' \bowtie T \bowtie S' \bowtie U).
\]
So the left route unfolds as
\[
  \rho_U
  \bigl(
    \operatorname{ap}(x_S, j \mapsto \bar a_j)
  \bigr)
  =
  (L^\sharp)_\varnothing(\rho(x_S))
  =
  (L^\sharp)_\varnothing
  \bigl(\sigma_{T \bowtie S' \ext \beta}(\kappa(x_S))\bigr).
\]

Now unfold the right-hand side.  Here \(\kappa\) fires immediately at the head
\(x_S\).  The filler being fired is \(\kappa(x_S)\), and we already recorded
that
\[
  \kappa(x_S)
  \in
  \E(P \bowtie D \bowtie T \bowtie S' \ext \beta).
\]
Define the pointwise action of \(\kappa\) on the original argument family by
\[
  (\kappa_U \cdot a)(j)
  =
  \kappa_{U_j}(a(j)).
\]
This is well-typed because
\[
  \kappa_{U_j} :
  \E(P \bowtie D \bowtie T \bowtie S \bowtie U_j)
  \to
  \E(P \bowtie D \bowtie T \bowtie S' \bowtie U_j),
\]
and \(a(j)\) is in the domain.  Hence
\[
  (\kappa_U \cdot a)(j)
  \in
  \E(P \bowtie D \bowtie T \bowtie S' \bowtie U_j).
\]
Its sharp is the singleton substitution
\[
  (\kappa_U \cdot a)^\sharp :
  \langle\beta\rangle
  \leadsto
  P \bowtie D \bowtie T \bowtie S' \bowtie U
\]
defined by
\[
  (\kappa_U \cdot a)^\sharp(\mathsf{here}(j))
  =
  \kappa_{U_j}(a(j)).
\]
The action at empty passive depth has type
\[
  ((\kappa_U \cdot a)^\sharp)_\varnothing :
  \E(P \bowtie D \bowtie T \bowtie S' \ext \beta)
  \to
  \E(P \bowtie D \bowtie T \bowtie S' \bowtie U).
\]
Its input \(\kappa(x_S)\) has exactly this domain type, so
\[
  ((\kappa_U \cdot a)^\sharp)_\varnothing(\kappa(x_S))
  \in
  \E(P \bowtie D \bowtie T \bowtie S' \bowtie U).
\]
Thus
\[
  \kappa_U(\operatorname{ap}(x_S,a))
  =
  ((\kappa_U \cdot a)^\sharp)_\varnothing(\kappa(x_S)).
\]
Applying \(\sigma_{T \bowtie S' \bowtie U}\) is now well-typed, since
\[
  \sigma_{T \bowtie S' \bowtie U} :
  \E(P \bowtie D \bowtie T \bowtie S' \bowtie U)
  \to
  \E(P \bowtie D' \bowtie T \bowtie S' \bowtie U).
\]
So the right route is
\[
  \sigma_{T \bowtie S' \bowtie U}
  \left(
    ((\kappa_U \cdot a)^\sharp)_\varnothing(\kappa(x_S))
  \right)
  \in
  \E(P \bowtie D' \bowtie T \bowtie S' \bowtie U).
\]
In Lean, the singleton substitution \((\kappa_U \cdot a)^\sharp\) is
\texttt{Subst.actPointwise kappa upsilon args}.

At this point the left route is
\[
  (L^\sharp)_\varnothing
  \bigl(\sigma_{T \bowtie S' \ext \beta}(\kappa(x_S))\bigr),
\]
and the right route is
\[
  \sigma_{T \bowtie S' \bowtie U}
  \left(
    ((\kappa_U \cdot a)^\sharp)_\varnothing(\kappa(x_S))
  \right).
\]

The recursive call that compares these is \texttt{diamondAt} applied to the
selected filler \(\kappa(x_S)\).  Its parameters are
\[
  T_{\mathrm{new}} = T \bowtie S',
  \qquad
  S_{\mathrm{new}}=\langle\beta\rangle,
  \qquad
  S'_{\mathrm{new}}=U,
  \qquad
  U_{\mathrm{new}}=\varnothing,
\]
with expression
\[
  \kappa(x_S)
  \in
  \E(P \bowtie D \bowtie T_{\mathrm{new}}
    \bowtie S_{\mathrm{new}}).
\]
The substitution supplied to this recursive call is
\[
  (\kappa_U \cdot a)^\sharp :
  S_{\mathrm{new}}
  \leadsto
  P \bowtie D \bowtie T_{\mathrm{new}} \bowtie S'_{\mathrm{new}},
\]
which is the same type as
\[
  \langle\beta\rangle
  \leadsto
  P \bowtie D \bowtie T \bowtie S' \bowtie U.
\]
In Lean, this is the call
\begin{quote}
\small\ttfamily
diamondAt (tau := tau bowtie dst) (src := singleton beta)\\
\hspace*{2em}(dst := Tele.ofList upsilon) sigma\\
\hspace*{2em}(upsilon := []) ... (Subst.actPointwise kappa upsilon args)\\
\hspace*{2em}(e := kappa xsrc)
\end{quote}
where \texttt{Subst.actPointwise kappa upsilon args} denotes
\((\kappa_U \cdot a)^\sharp\).

Let us now unfold what this recursive call says before simplifying its
notation.  The general diamond theorem, applied with
\[
  T_{\mathrm{new}}=T\bowtie S',
  \quad
  S_{\mathrm{new}}=\langle\beta\rangle,
  \quad
  S'_{\mathrm{new}}=U,
  \quad
  U_{\mathrm{new}}=\varnothing,
\]
says:
\[
\begin{aligned}
&
  \left(
    \pcomp{\sigma}{T_{\mathrm{new}}\bowtie S'_{\mathrm{new}}}
      {(\kappa_U \cdot a)^\sharp}
  \right)_{\varnothing}
  \Bigl(
    \sigma_{T_{\mathrm{new}}\bowtie S_{\mathrm{new}}}
      (\kappa(x_S))
  \Bigr)
\\
&\qquad =
  \sigma_{T_{\mathrm{new}}\bowtie S'_{\mathrm{new}}}
  \left(
    ((\kappa_U \cdot a)^\sharp)_{\varnothing}(\kappa(x_S))
  \right).
\end{aligned}
\]
Now substitute the definitions of the new parameters.  Since
\[
  T_{\mathrm{new}}\bowtie S_{\mathrm{new}}
  =
  T \bowtie S' \ext \beta,
  \qquad
  T_{\mathrm{new}}\bowtie S'_{\mathrm{new}}
  =
  T \bowtie S' \bowtie U,
\]
the recursive equation becomes
\[
\begin{aligned}
&
  \left(
    \pcomp{\sigma}{T \bowtie S' \bowtie U}
      {(\kappa_U \cdot a)^\sharp}
  \right)_{\varnothing}
  \bigl(
    \sigma_{T \bowtie S' \ext \beta}(\kappa(x_S))
  \bigr)
\\
&\qquad =
  \sigma_{T \bowtie S' \bowtie U}
  \left(
    ((\kappa_U \cdot a)^\sharp)_\varnothing(\kappa(x_S))
  \right).
\end{aligned}
\]
The right-hand side is already the right-hand side we want.  The left-hand side
still contains the postcomposition
\[
  \pcomp{\sigma}{T \bowtie S' \bowtie U}{(\kappa_U \cdot a)^\sharp}.
\]
This is a singleton-domain substitution
\[
  \langle\beta\rangle
  \leadsto
  P \bowtie D' \bowtie T \bowtie S' \bowtie U.
\]
Its filler at \(\mathsf{here}(j)\) is, by definition of postcomposition,
\[
\begin{aligned}
&
  \left(
    \pcomp{\sigma}{T \bowtie S' \bowtie U}
      {(\kappa_U \cdot a)^\sharp}
  \right)(\mathsf{here}(j))
\\
&\qquad =
  \sigma_{T \bowtie S' \bowtie U_j}
  \left(
    (\kappa_U \cdot a)^\sharp(\mathsf{here}(j))
  \right)
\\
&\qquad =
  \sigma_{T \bowtie S' \bowtie U_j}
  \bigl(\kappa_{U_j}(a(j))\bigr).
\end{aligned}
\]
This is exactly the expression that we call \(R_j\) below.  Thus the
postcomposed singleton substitution is \(R^\sharp\).  Consequently its action
at empty passive depth is
\[
  (R^\sharp)_\varnothing :
  \E(P \bowtie D' \bowtie T \bowtie S' \ext \beta)
  \to
  \E(P \bowtie D' \bowtie T \bowtie S' \bowtie U).
\]
The input
\[
  \sigma_{T \bowtie S' \ext \beta}(\kappa(x_S))
\]
has precisely this domain type, because \(\kappa(x_S)\) lives in
\(\E(P \bowtie D \bowtie T \bowtie S' \ext \beta)\).  Therefore the left-hand
side of the recursive diamond is exactly
\[
  (R^\sharp)_\varnothing
  \bigl(\sigma_{T \bowtie S' \ext \beta}(\kappa(x_S))\bigr).
\]

So, after naming this postcomposed singleton substitution \(R^\sharp\), the
recursive theorem says exactly that
\[
  (R^\sharp)_\varnothing
  \bigl(\sigma_{T \bowtie S' \ext \beta}(\kappa(x_S))\bigr)
  =
  \sigma_{T \bowtie S' \bowtie U}
  \left(
    ((\kappa_U \cdot a)^\sharp)_\varnothing(\kappa(x_S))
  \right),
\]
where
\[
  R_j =
  \sigma_{T \bowtie S' \bowtie U_j}
  \bigl(\kappa_{U_j}(a(j))\bigr).
\]
This is well-typed because \(\kappa_{U_j}(a(j))\) lives in
\(\E(P \bowtie D \bowtie T \bowtie S' \bowtie U_j)\), so
\[
  R_j
  \in
  \E(P \bowtie D' \bowtie T \bowtie S' \bowtie U_j).
\]
Similarly, \(R^\sharp\) is the singleton substitution generated by
\(j \mapsto R_j\):
\[
  R^\sharp :
  \langle\beta\rangle
  \leadsto
  P \bowtie D' \bowtie T \bowtie S' \bowtie U.
\]
The only remaining mismatch is that our unfolded left route contains \(L_j\),
not \(R_j\).  But the recursive hypotheses already computed for the original
arguments \(a(j)\) say precisely that.  Fix one binder
\[
  j : \B(\beta).
\]
The \(j\)-th argument is the smaller expression
\[
  a(j) \in \E(P \bowtie D \bowtie T \bowtie S \bowtie U_j).
\]
The ordinary argument-recursive call of \texttt{diamondAt} is made with the
same substitutions \(\sigma\) and \(\kappa\), the same telescopes
\[
  P,\quad D,\quad D',\quad T,\quad S,\quad S',
\]
but with the under-telescope changed from \(U\) to
\[
  U_j = U \ext \arity(j),
\]
and with expression \(a(j)\).  In Lean this family of recursive calls is
\begin{quote}
\small\ttfamily
hargs j := diamondAt (upsilon := j.arity :: upsilon)\\
\hspace*{2em}sigma hTauSrc hTauDst srcTarget dstTarget kappa (args j)
\end{quote}
Mathematically, this instance of the theorem says
\[
  \rho_{U_j}
  \bigl(
    \sigma_{T \bowtie S \bowtie U_j}(a(j))
  \bigr)
  =
  \sigma_{T \bowtie S' \bowtie U_j}
  \bigl(
    \kappa_{U_j}(a(j))
  \bigr).
\]
The left-hand side is exactly \(L_j\), because
\[
  \bar a_j = \sigma_{T \bowtie S \bowtie U_j}(a(j))
  \quad\text{and}\quad
  L_j = \rho_{U_j}(\bar a_j).
\]
The right-hand side is exactly \(R_j\), by the definition of \(R_j\).  Hence
\(\texttt{hargs j}\) is precisely the equality
\[
  L_j = R_j
\]
for this \(j\).

The equality named \texttt{hSubst} in Lean is the extensional equality between
the two singleton-domain substitutions generated by the two argument families:
\[
  L^\sharp = R^\sharp.
\]
Both substitutions have domain \(\langle\beta\rangle\) and codomain
\[
  P \bowtie D' \bowtie T \bowtie S' \bowtie U.
\]
To prove two such substitutions equal, it is enough to check their value at
every singleton slot.  There is only one nonempty form of such a slot:
\[
  \mathsf{here}(j), \qquad j : \B(\beta).
\]
At that slot, the desired equality is exactly
\[
  L^\sharp(\mathsf{here}(j)) = R^\sharp(\mathsf{here}(j)),
\]
which unfolds to \(L_j = R_j\), i.e. exactly \(\texttt{hargs j}\).  This is why
the Lean proof of \texttt{hSubst} has one real case, \(\mathsf{here}(j)\), and
that case is closed by \texttt{hargs j}.

This recursive call is terminating because it decreases the source side of the
diamond: instead of proving the theorem for all of \(S\), it proves it for the
single selected \(S\)-slot \(x_S\).

\paragraph{Head in \(T\).}
This head is passive for both substitutions.  Both sides preserve it and
recurse on the arguments.

Concretely, suppose
\[
  e = \operatorname{ap}(x_T,a),
\]
where \(x_T : T \ni \beta\).  This head lies in the ambient passive telescope
between the \(D\)-part and the \(S\)-part.  It is not a variable that
\(\sigma\) may replace, because \(\sigma\) only acts on \(D\).  It is also not a
variable that \(\kappa\) may replace, because \(\kappa\) only acts on \(S\).
Thus both routes through the diamond rebuild the same \(T\)-head.

For each binder argument \(j \in \B(\beta)\), set
\[
  L_j =
  (\pcomp{\sigma}{T \bowtie S'}{\kappa})_{U \ext \arity(j)}
  \bigl(
    \sigma_{T \bowtie S \bowtie U \ext \arity(j)}(a(j))
  \bigr)
\]
and
\[
  R_j =
  \sigma_{T \bowtie S' \bowtie U \ext \arity(j)}
  \bigl(
    \kappa_{U \ext \arity(j)}(a(j))
  \bigr).
\]
The recursive hypothesis again gives \(L_j = R_j\) for every \(j\).  The
recursive call is exactly the same one as in the \(U\)-case: for each argument
\(j\), we apply the theorem to the subterm \(a(j)\), leaving
\(\sigma,\kappa,T,S,S'\) and all coherence data unchanged, and replacing the
under-telescope \(U\) by
\[
  U_j = U \ext \arity(j).
\]
In Lean, again with ASCII names for readability, this is
\begin{quote}
\small\ttfamily
diamondAt (upsilon := j.arity :: upsilon)\\
\hspace*{2em}sigma hTauSrc hTauDst srcTarget dstTarget kappa (args j)
\end{quote}
Therefore
the \(T\)-case reduces to
\[
  \operatorname{ap}(x_T, j \mapsto L_j)
  =
  \operatorname{ap}(x_T, j \mapsto R_j).
\]
The Lean proof is longer than this sentence because it must show that the
formal embeddings of the same \(T\)-head through
\(P \bowtie D \bowtie T \bowtie S\) and
\(P \bowtie D \bowtie T \bowtie S'\) are the intended ones.  Once this
reassociation bookkeeping is done, the branch is just congruence plus the
recursive equalities.

\paragraph{Head in \(D\).}
This is an active variable for \(\sigma\).  Firing \(\sigma\) creates a
one-binder instantiation from the arguments of the application.  This is the
only branch where the companion theorem for one-binder instantiation is needed.

\paragraph{Head in \(P\).}
This is part of the untouched prefix.  Both substitutions preserve it and
recurse on the arguments.

Here an application has the form
\[
  e = \operatorname{ap}(x_P,a),
\]
with \(x_P : P \ni \beta\).  The prefix \(P\) is shared background context:
\(\sigma\) preserves it while replacing only \(D\)-variables, and \(\kappa\)
preserves it while replacing only \(S\)-variables.  Consequently the head
\(x_P\) survives unchanged along both sides of the square.

As before, for each binder argument \(j\) we compare the two transformed
subterms
\[
  L_j =
  (\pcomp{\sigma}{T \bowtie S'}{\kappa})_{U \ext \arity(j)}
  \bigl(
    \sigma_{T \bowtie S \bowtie U \ext \arity(j)}(a(j))
  \bigr),
  \qquad
  R_j =
  \sigma_{T \bowtie S' \bowtie U \ext \arity(j)}
  \bigl(
    \kappa_{U \ext \arity(j)}(a(j))
  \bigr).
\]
The induction hypothesis gives \(L_j = R_j\).  Concretely, for each \(j\) the
recursive call is the theorem applied to \(a(j)\) at the enlarged passive
under-telescope
\[
  U_j = U \ext \arity(j),
\]
with the same substitutions \(\sigma\) and \(\kappa\), the same ambient
telescope \(T\), the same source and destination \(S,S'\), and the same
coherence witnesses.  In Lean, using ASCII names for readability, it is the
same call
\begin{quote}
\small\ttfamily
diamondAt (upsilon := j.arity :: upsilon)\\
\hspace*{2em}sigma hTauSrc hTauDst srcTarget dstTarget kappa (args j)
\end{quote}
Hence
\[
  \operatorname{ap}(x_P, j \mapsto L_j)
  =
  \operatorname{ap}(x_P, j \mapsto R_j).
\]
The only formal work in this case is showing that the prefix injection for
\(x_P\) is the same after passing through the source and destination
parenthesizations.  Mathematically, nothing happens to the head.

\section*{Why the formal statement has coherence data}

The mathematical equation above suppresses associativity bookkeeping for
\(\bowtie\).  Lean cannot suppress it completely.  The formal theorem therefore
carries witnesses saying that the decompositions
\[
  T \bowtie S
  \qquad\text{and}\qquad
  T \bowtie S'
\]
are proper and that injections into larger contexts compute with the intended
parenthesization.  These witnesses do not change the mathematical content of
the theorem.  They only ensure that the formal proof can recognize, for example,
that a head from \(S\) remains a head from \(S\) after it is embedded through
\[
  P \bowtie D \bowtie T \bowtie S.
\]

\section*{Role in substitution composition}

The public composition theorem says that acting by a composite substitution is
the same as acting by one substitution and then the other.  The difficult case
is when the first substitution fires on an application head.  Firing creates a
fresh one-binder instantiation, and the remaining substitution must commute
past that instantiation.

The theorem described here is the general interchange principle that makes that
step possible.  In short:
\[
  \text{hereditary substitution respects substitution composition.}
\]

\end{document}
